Effectful Semantics for Substructural SSA

### 2024-02-24 toplas/paper.tex
Tap

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
\setcopyright{acmcopyright}
\copyrightyear{2023}
\acmYear{2023}
\acmDOI{XXXXXXX.XXXXXXX}

@@

\begin{itemize}
\item We give a type system for SSA-style programs, which has a full
substitution theorem and equational theory for both expressions and control-flow
graphs. This language includes substructural types, both for controlling effects
and tracking ownership in values. 
\item We give an axiomatic semantics for our language, by showing that it has a
compositional intepretation in any distributive effectful category (a
substructural variant of a Freyd category) with an Elgot structure. We use this
structure to give generic proofs of the soundness of substitution and rewriting
properties.
\item Our abstract semantics justifies a variety of powerful optimisation
techniques, including inlining, loop fusion, hoisting, and strength reduction. 
\item We show how Elgot monads give rise to instances of our categorical
semantics. We use this to derive semantics from a general trace monad, and show
that various monad transformers such as state transformers preserve the needed
Elgot structure. 
\item We use these tools to demonstrate the existence of a variety of concrete
models satisfying these categorical axioms. Starting first from simple languages
such as state plus printing, we also show how more challenging semantics such as
the TSO semantics for weak memory also fit into our framework. This thus gives
rise to an SSA-based IR with support for weak memory operations, which is fully
semantically-justified.
% \item Our weak memory semantics and our results about Elgot monads are
% formalised in the Lean theorem prover.
\end{itemize}

@@

\begin{figure}
  \centering
  \begin{subfigure}[t]{0.33\textwidth}
    \begin{small}
    \begin{BVerbatim}[baseline=t, gobble=3]
    'entry:
      m = 0
      n = 1
      brz i 'exit 'loop
    'loop:
      t = add m n
      m = n
      n = t
      i = sub i 1
      brz i 'exit 'loop
    'exit:
      ret m
    \end{BVerbatim}
    \end{small}
    \caption{3-address code}
    \label{fig:3-addr}
  \end{subfigure}
  \begin{subfigure}[t]{0.32\textwidth}
    \begin{small}
    \begin{BVerbatim}[baseline=t, gobble=3]
    'entry:
      m0 = 0
      n0 = 1
      brz i
        'exit (m0) 
        'loop (i, m0, n0)
    'loop(i0, m1, n1):
      m2 = n1
      n2 = add m1 n1
      i1 = sub i0 1
      brz i1 
        'exit (m2) 
        'loop (i1, m2, n2)
    'exit(m3):
      ret m3
    \end{BVerbatim}
    \end{small}
    \caption{Basic blocks with arguments}
    \label{fig:bb-arg}
  \end{subfigure}
  \begin{subfigure}[t]{0.33\textwidth}
    \begin{small}
    \begin{BVerbatim}[baseline=t, gobble=3]
    let m0 = 0;
    let n1 = 1;
    if eq i 0 {
      br 'exit (m0)
    } else {
      br 'loop (i, m0, n0)
    } where
    'loop(i0, m1, n1) =>
      let m2 = n1;
      let n2 = add m1 n1;
      let i1 = sub i0 1;
      if eq i1 0 {
        br 'exit (m2)
      } else {
        br 'loop (i1, m2, n2)
      }
    'exit(m3) => ret m3
    \end{BVerbatim}
    \end{small}
    \caption{Our syntax}
    \label{fig:our-syn}
  \end{subfigure}
  \caption{A simple program for computing the \texttt{i}th Fibonacci number,
  represented as 3-address code, basic-blocks with arguments, and in our SSA
  syntax} 
  \Description{}
  \label{fig:ssa-examples}
\end{figure}

@@

  \item \textit{Blocks} \(s, t\), which are typed via the judgement
  \(\haslb{\Gamma}{t}{\ms{L}}\), which says that the block \(t\) is well-formed
  with respect to the context \(\Gamma\) and only branches to
  ("\textit{targets}") labels in \(\ms{L}\). The rules for this judgement are
  given in Figure \ref{fig:ssa-block-typing}.

\item \textit{Control-flow graphs} or \textit{CFG}s \(L\), which are typed via
  the judgement \(\lhaslb{\ms{L}}{L}{\ms{K}}\), which says that \(L\) defines a
  control-flow graph with input labels \(\ms{L}\) and output labels \(\ms{K}\);
  recursive control flow is allowed. The rules for this judgement are given in
  Figure \ref{fig:ssa-block-typing}.
\end{itemize}
The formal grammar for our syntax is given in Figure \ref{fig:ssa-grammar} in
the appendix, while the typing judgements themselves are given in Figure
\ref{fig:ssa-judgements}.

>>>>

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
\setcopyright{acmcopyright}
\copyrightyear{2024}
\acmYear{2024}
\acmDOI{XXXXXXX.XXXXXXX}

@@

\begin{itemize}
\item We give a type system for SSA-style programs, which has a full
substitution theorem and equational theory for both expressions and control-flow
graphs. This language includes substructural types, both for controlling effects
and tracking ownership in values. 
\item We give an axiomatic semantics for our language, by showing that it has a
compositional intepretation in any distributive effectful category (a
substructural variant of a Freyd category) with an Elgot structure. We use this
structure to give generic proofs of the soundness of substitution and rewriting
properties.
\item We show how our syntax quotiented by our equational theory is an initial model
of our semantics.
\item Our abstract semantics justifies a variety of powerful optimisation
techniques, including inlining, loop fusion, hoisting, and strength reduction. 
\item We show how Elgot monads give rise to instances of our categorical
semantics. We use this to derive semantics from a general trace monad, and show
that various monad transformers such as state transformers preserve the needed
Elgot structure. 
\item We use these tools to demonstrate the existence of a variety of concrete
models satisfying these categorical axioms. Starting first from simple languages
such as state plus printing, we also show how more challenging semantics such as
the TSO semantics for weak memory also fit into our framework. This thus gives
rise to an SSA-based IR with support for weak memory operations, which is fully
semantically-justified.
% \item Our weak memory semantics and our results about Elgot monads are
% formalised in the Lean theorem prover.
\end{itemize}

@@

\begin{figure}
  \centering
  \begin{subfigure}[t]{0.33\textwidth}
    \begin{small}
    \begin{BVerbatim}[baseline=t, gobble=3]
    ^entry:
      m = 0
      n = 1
      brz i 'exit 'loop
    ^loop:
      t = add m n
      m = n
      n = t
      i = sub i 1
      brz i 'exit 'loop
    ^exit:
      ret m
    \end{BVerbatim}
    \end{small}
    \caption{3-address code}
    \label{fig:3-addr}
  \end{subfigure}
  \begin{subfigure}[t]{0.32\textwidth}
    \begin{small}
    \begin{BVerbatim}[baseline=t, gobble=3]
    ^entry:
      m0 = 0
      n0 = 1
      brz i
        ^exit (m0) 
        ^loop (i, m0, n0)
    ^loop(i0, m1, n1):
      m2 = n1
      n2 = add m1 n1
      i1 = sub i0 1
      brz i1 
        ^exit (m2) 
        ^loop (i1, m2, n2)
    ^exit(m3):
      ret m3
    \end{BVerbatim}
    \end{small}
    \caption{Basic blocks with arguments}
    \label{fig:bb-arg}
  \end{subfigure}
  \begin{subfigure}[t]{0.33\textwidth}
    \begin{small}
    \begin{BVerbatim}[baseline=t, gobble=3]
    let m0 = 0;
    let n1 = 1;
    if eq i 0 {
      br ^exit (m0)
    } else {
      br ^loop (i, m0, n0)
    } where
    ^loop(i0, m1, n1) =>
      let m2 = n1;
      let n2 = add m1 n1;
      let i1 = sub i0 1;
      if eq i1 0 {
        br ^exit (m2)
      } else {
        br ^loop (i1, m2, n2)
      }
    ^exit(m3) => ret m3
    \end{BVerbatim}
    \end{small}
    \caption{Our syntax}
    \label{fig:our-syn}
  \end{subfigure}
  \caption{A simple program for computing the \texttt{i}th Fibonacci number,
  represented as 3-address code, basic-blocks with arguments, and in our SSA
  syntax} 
  \Description{}
  \label{fig:ssa-examples}
\end{figure}

@@

  \item \textit{Bodies} \(b\), which are typed via the judgement
  \(\hasty{\Gamma}{p}{b}{\Delta}{}\), which says that, given variables
  \(\Gamma\) live on entry, the variables \(\Delta\) are live on exit. The rules
  for this judgement are given in Figure \ref{fig:ssa-block-typing}.

  \item \textit{Terminators} \(s, t\), which are typed via the judgement
  \(\haslb{\Gamma}{t}{\ms{L}}\), which says that the terminator \(t\) is
  well-formed with respect to the context \(\Gamma\) and only branches to
  ("\textit{targets}") labels in \(\ms{L}\). The rules for this judgement are
  given in Figure \ref{fig:ssa-block-typing}.

  \item \textit{Basic blocks} \(\beta\), which are typed via the judgement
  \(\haslb{\Gamma}{\beta}{\ms{L}}\)...

  \item \textit{Control-flow graphs} or \textit{CFG}s \(L\), which are typed via
    the judgement \(\lhaslb{\ms{L}}{L}{\ms{K}}\), which says that \(L\) defines a
    control-flow graph with input labels \(\ms{L}\) and output labels \(\ms{K}\);
    recursive control flow is allowed. The rules for this judgement are given in
    Figure \ref{fig:ssa-block-typing}.

  \item \textit{Regions} ...
\end{itemize}
The formal grammar for our syntax is given in Figure \ref{fig:ssa-grammar} in
the appendix, while the typing judgements themselves are given in Figure
\ref{fig:ssa-judgements}.
